ISI Control in stochastic LIF models

### 2013-08-22 log.tex
Initial commit

\usepackage{cleveref}
\usepackage{cancel}

@@

\title{Stochastic Optimal Control - with a view towards spike control in neurons
\\
{\small Advanced Comprehensive Exam in Mathematics}
\\
{\small Dept. of Math - University of Ottawa}
}
\author{Alexandre Iolov 
$<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$}

\date{January 10, 2013}

@@

\abstract{We review main concepts from stochastic optimal control theory,
including stochastic differential equations, dynamic programing and Pontryagin's
Maximum Principle. We then discuss an optimal control problem in a simple
mathematical model of a neuron where the feedback is minimal and we have a
quasi-open-loop control which necessitates the use of a Maximum Principle for
the system's transition density.}

@@

\section{Overview}
Stochastic Optimal Control in continuous time deals with finding
the input which optimizes the performance of a continuous-time dynamical systems
that contain a significant random component in their evolutions.

Here we present a brief overview of the subject starting from the basic notions
of Stochastic Differential Equations (SDEs), 
and then introducing the main tools of Optimal Control, namely the
Hamilton-Jacobi-Bellman equation of Dynamic Programming and the state-adjoint
equations arising from Pontryagin's Maximum Principle. We will not state
proofs, although we sometimes provide heuristic derivations of the main
results.

If I had to learn the material all over again, I would proceed my readings as
follows: First I would read the online notes by Evans on both
SDEs,\cite{Evansa}, and Optimal Control, \cite{Evansb}. I would then read
Jacobs' book, \cite{Jacobs}, especially for the chapter on the Fokker-Planck
equation (ch 7) and then Oksendal \cite{Oksendal2007} for a
more thorough mathematical treatment of SDEs. Having familiarized myself with SDE's I would
return to Optimal Control in the Deterministic Case, starting with \cite{Kirk2004} for the
finite dimensional deterministic case (ODEs) and then \cite{Fleming1975} for the
extension to the stochastic case. I would keep the following for reference \cite{Krylov2008},
\cite{Hanson2007} and finally \cite{Fattorini1999} for the
deterministic infinite-dimensional (PDEs) case.

I have not used any particularly sophisticated numerical
methods, but, for example, Hanson \cite{Hanson2007} discusses both the
Euler-Maruyama scheme for simulating SDEs and the Crank-Nicholson Method for
solving PDEs which is all I have used in my codes.

All of the theoretical results presented in the sequel will be from
the above references.

\section{Stochastic Differential Equations}
In view of our ultimate goals, we will restrict ourselves to  stochastic
processes whose sample paths have continuous paths, i.e.\ to SDEs driven by
Brownian motion.

Since the reader is more likely well familiar with the material in this section,
we will not provide proofs, but only state the results with a view towards
establishing the notation for the sequel.

We will assume that the reader is familiar with the following concepts:
\begin{itemize} 
  \item a probability space, $\{\O, \sAlg, P\}$ consisting of a
  probability space, a sigma-algebra and a probability measure
  \item a continuous-time stochastic process, $X_t$
  \item a filtration $\Fil(t)$ and in particular the filtration generated by a
  stochastic process.
\end{itemize}

\subsection{The Wiener Process and the Ito Integral}
The Wiener Process is the fundamental building block of the stochastic calculus,
it is often called Brownian Motion and we will denote it $\{W_t\}_{t\geq 0}$. It
satisfies:
\begin{defn}Wiener Process, $W_t$:
\begin{enumerate}
  \item $W_0 = 0$
  \item $W_t - W_s = N(0, |t-s|)$ , i.e. normally distributed with mean 0,
  variance $|t-s|$)
  \item $\forall \{t_i\}_1^N, \quad \{W_{t_i} - W_{t_{i-1}} \}_2^N \sim$
  independent, i.e $W_t$ has independent increments
\end{enumerate}
\end{defn}
The fact that the finite incremental distributions suffice to specify a unique
continuous-time stochastic process is known as Kolmogorov's extension theorem.
\begin{ex}[from \cite{Evansa} \#21.] Inverted Wiener Process:
\\
Let $W_t$ be a 1-d Wiener Process and define:
\def \Wbar {{\bar{W} }}
\begin{equation*}
\Wbar(t) := \begin{cases}
t W(1/t) & t >0
\\
0 & t = 0
\end{cases}
\end{equation*}
Show $\Wbar(t)- \Wbar(s) \sim N(0, t-s)$.

\begin{proof}[Soln]
\begin{align*}
\Wbar(t) - \Wbar(s) =& t W(1/t) - s W(1/s), \, t>s
\\
=& (t-s)W(1/t) + s W(1/t) - sW(1/s) & \text{recall:} 1/s > 1/t
\\
=& (t-s)N(0, 1/t) - sN(0, 1/s -1/t) & \text{by indnt increments}
\\
=& N\left(0, \frac{(t-s)^2}t + \frac{s^2 }{ 1/s -1/t} \right) & \text{
addition of indnt normal RVs.}
\end{align*}
With some algebra, the variance in the last expression reduces to $t-s$
\end{proof}
\end{ex}

Now, we collect a few more relevant properties of $W_t$:
\begin{enumerate}
  \item The sample paths $W_{[0,\infty)}(\o)$
are almost surely (a.s.) continuous and are in fact Holder continuous for any
exponent $\g < 1/2$
\item The sample paths $W_{[0,\infty)}(\o)$ are nowhere differentiable
\item $W_t$ is a Markov process: $\Prob[W_t \in B | \s(W_{s' \leq s })] =
\Prob[W_t \in B \,| \,W_s]$, for any Borel set, $B \subset \R$, where $\s(W_{s'
\leq s })$ is the filtration generated by the $W_t$
\end{enumerate}

We now turn to defining stochastic integrals based on the Wiener Process.

\begin{defn} Progressively Measurable Functions:

Let $\Fil_t$ be the filtration generated by the Wiener Process.

Let $X_t$ be a stochastic process which is $\Fil_t$-measurable $\forall t$ and
which is jointly measurable in $(t,\o)$. We call such an $X_t$
\emph{progressively measurable}
\end{defn}

\begin{defn} $\Ltwopm, \Lonepm$

We define $\Ltwopm[0,T]$ as the space of all progressively measurable
$X$ such that
\begin{equation*}
\Exp \left[ \int_{[0,T]} X_t^2 \intd{t} \right] < \infty
\end{equation*}

Similarly, we define $\Lonepm[0,T]$ as the space of all progressively measurable
$X$ such that
\begin{equation*}
\Exp \left[ \int_{[0,T]} X_t \intd{t} \right] < \infty
\end{equation*} 
\end{defn}

$\Ltwopm$ will be the class of functions for which the Ito integral is
well-defined. 
\begin{defn} Ito Integral:
\label{defn:ito_integral}

Let $P^n := {a = t^n_1 \ldots t^n_{m_n} = b}$ be a partition of the interval
$[a,b] \subset [0, \infty)$. Let $|P_n| = \sup_i|t_i - t_{i-1}|$. 
Take $|P_n| \rightarrow_n 0 $ and consider an $ X_t \in \Ltwopm[a,b]$,

then
\begin{equation}
\int_{[a,b]} X_t \intd{W} := \lim_{n \rightarrow \infty}  
\sum_{i=1}^{m_n} X_{t_i}\left( W(t_{i+1}) - W(t_{i})\right)
\end{equation}

\end{defn}

To be precise, our definition is actually a theorem, and the real
definition is one that uses step functions and passes to the limit.  Also we
will write the limits of integration $\int_0^T \cdot  \intd{W}$ or $\int_{[0,T]}
\cdot  \intd{W}$ interchangeably.
% To justify def'n \ref{defn:ito_integral}, we recall:
% \begin{lemma}[Quadratic Variation]
% \begin{equation}
% \lim_{n \rightarrow \infty} \sum_{i=1}^{m_n} \left( W(t_i) - W(t_{i-1})\right)^2
% = b-a \quad \text{in }  L^2(\O)
% \end{equation}
% \end{lemma}
% The above can be read, heuristically as $\lim \sum (\Delta W)^2 \rightarrow \sum
% \Delta t$, which is the meaning behind the colloquial $\intd{W} =
% \sqrt{dt}$.

Again, we state without proof a few interesting properties of $\int X \intd{W}$:
\begin{thm} Ito Integral Properties

\begin{enumerate}
  \item $\Exp[\int_0^T X \intd{W} ] = 0$ 
  \item $\Exp[\left(\int_0^T X \intd{W}\right)^2 ] = \Exp[\int_0^T X^2
  \intd{t}]$
  \item $I(t) = \int_0^t X \intd{W} $ is a martingale 
  \item $I(t) = \int_0^t X \intd{W} $ has continuous sample paths
\end{enumerate}
\end{thm}

\subsection{Ito SDEs and Ito's Lemma}
\begin{defn} an Ito SDE:

We write
\begin{equation}
dX_t = Fdt + G dW
\end{equation}
on $0 \leq t \leq T$, if $X$ is a real-valued stochastic process satisfying:
\begin{equation*}
X(r) = X(s) + \int_{[s,r]} F \intd{t} + \int_{[s,r]} G \intd{W}
\quad 0\leq s \leq r \leq T
\end{equation*}
for some $F \in \Lonepm[0,T]$ and $G \in \Ltwopm[0,T]$
\end{defn}

We are now ready to present the celebrated Ito Lemma which is the chain-rule of
stochastic calculus:

\begin{thm}[Ito Lemma] Suppose $X$ satisfies the stochastic differential $dX_t =
Fdt + G dW$ as above and take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$.

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)
\intd{t} + \left(   \di_x u\cdot G  \right)\intd{W}
$$
\end{thm}

% We now present two examples, one fundamental and one more involved:
% \begin{ex}[pg 67 in \cite{Evansa}] The 'Stochastic' Exponential:
% 
% Let $Y = \exp(\l W - \l^2 t / 2)$. In the notation of the above Thm, we have:
% $dX = dW, F=0, G=1$ and $u(x,t) = \exp(\l x - \l^2 t / 2)$.
% 
% then
% \begin{align*}
% dY =& 
% \left( \di_t u + \di_x u \cdot F + \di^2_x u \cdot \frac{G^2}2 \right)\intd{t} +
% \left(   \di_x u \cdot G  \right)\intd{W}
% \\
% =&
% \left( -\frac{\l^2 t}{2}\exp(\l x - \l^2 t / 2) + \l^2\exp(\l x - \l^2 t / 2)  \right)\intd{t}
% \\ 
% &+ \left(   \l \exp(\l x - \l^2 t / 2) \right)\intd{W}  
% \\
% =&  \l \left( \exp(\l x - \l^2 t / 2) \right)\intd{W}  
% \\
% =& \l Y dW
% \end{align*}
% \end{ex}
% 
% \begin{ex}[pg. 67 in \cite{Evansa}] Hermite Polynomials:
% 
% Let $h_n(x,t) = \frac{(-t)^n}{n!} e^{x^2/2t} \di^n_x \left( e^{x^2/2t} 
% \right)$ be the $n$th Hermite polynomial
% \\
% then
% $$
% \int_0^t h_n(W, t) \intd{W} = h_{n+1}(W(t), t)) 
% $$
% i.e. in Ito calculus, the Hermite polynomials are like the power monomials
% $\frac{t^n}{n!}$ in the ordinary calculus.
% \begin{proof} Since
% $$
% d_\l^n [\exp(- \frac{(x-\l t)^2}{2t} ] |_{\l=0} =
%  (-t)^n d_x^n [ \exp(-\frac{x^2}{2t})] $$
% We will have:
% \begin{align*}
% d_\l^n [\exp(\l x - \l^2 t / 2) ] |_{\l=0} &= (-t)^n  e^{x^2/2t} d_x^n [
% \exp(-\frac{x^2}{2t})]
% \\
% &= n! h_n(x,t)
% \end{align*}
% Thus we can expand $\exp(\l x - \l^2 t / 2) $ in a Taylor series at $\l=0$ as:
% $$
% \exp(\l x - \l^2 t / 2) = \sum_n \l^n h_n(x,t)
% $$
% Now recall $Y$ from the previous example:
% \begin{equation*}
% Y(t) = \exp(\l W(t) - \frac{\l^2 t }2) = \sum_n \l^n h_n(W(t),t)
% \end{equation*}
% which satisfies:
% \begin{equation*}
% Y(t) = 1 + \l \int Y \intd{W}
% \end{equation*}
% Naturally we will plug the series expansion in the SDE to obtain:
% \begin{align*}
% \sum_{n=0}^{\infty} \l^n h_n(x,t) =&
% 1 + \l \int_0^t  \sum_{n=0}^{\infty} \l^n
% h_n(x,t)\intd{W}
% \\
% =& 1 + \sum_{n=1}^{\infty} \l^n\int_0^t 
% h_{n-1}(x,t)\intd{W} \quad {\textrm{bump $n$ up by one}}
% \end{align*}
% Equating the coefficients for $\l^n$ on both sides leads to the result
% \end{proof}
% \end{ex}

Finally, we are in a position to define an Ito SDE:

\begin{defn}[Ito SDE] Let
$F(x,t):\R\times [0,T]\ra \R,  G(x,t):\R\times [0,T]\ra \R,$ be given functions.
We say that a stochastic process $X$ satisfies:
\begin{equation}
dX =F dt + G dW
\end{equation} 
over $[0,T]$ if
\begin{enumerate}
  \item $X$ is progressively measurable wrt. $\Fil_t$
  \item $F(X,t) \in \Lonepm$
  \item $G(X,t) \in \Ltwopm$ 
  \item $X_t = X_0 + \int_0^t F(X_s, s) \intd{s} + \int_0^t G(X_s, s) \intd{W_s}$
\end{enumerate}
\end{defn}

% \begin{ex} The solution to 
% $$
% \begin{cases}
% dX = F(t)X dt + G(t) X dW
% \\
% X_0 = 1
% \end{cases}
% $$
% is 
% $$
% X_t = \exp\left( \int_0^t F(s) - \frac{G^2(s)}{2} \intd{s} + \int_0^t G
% \intd{W} \right) $$
% \begin{proof}
% Let $Y = \int_0^t f(s) - \frac{g^2(s)}{2} \intd{s} + \int_0^t g
% \intd{W} $
% and guess that $X = e^Y$
% then, using Ito's lemma:
% \begin{align*}
% dX =& (\di_y [e^y] dY + \frac{1}{2}\di^2_y[e^y]  g^2 dt
% \\ =&
% (e^y  f dt - e^y g^2 / 2) dt + e^y g dW + \frac{1}{2} e^y \cdot  g^2 dt
% \\ =& e^y \left( f dt +g dW \right)
% \\ =& X \left( f dt +g dW \right)
% \end{align*}
% \end{proof}
% \end{ex}

% \begin{ex}[Brownian Bridge, pg 80 in \cite{Evansa}] The solution of the SDE
% $$
% \begin{cases}
% dB = -\frac{B}{1-t} dt + dW
% \\
% B_0= 0
% \end{cases}
% $$
% is
% $$
% B(t) = (1-t) \int_0^t \frac{1}{1-s} dW
% $$
% \begin{proof}
% It is not clear how to apply Ito's lemma to this problem, so instead we just
% directly calculate the differential:
% \begin{align*}
% B(t^+) - B(t) =& (1-t^+) \int_0^{t^+} \frac{1}{1-s} dW - (1-t) \int_0^t
% \frac{1}{1-s} dW
% \\ =&
% (1-t^+) \int_t^{t^+}\frac{1}{1-s} dW  - (t^+-t) \int_0^t
% \frac{1}{1-s} dW
% \\ \approx&
% \int_t^{t^+} dW - \Delta t \frac{B(t)}{1-t}  
% \end{align*}
% \end{proof}
% \end{ex}
% Actually we can say a bit more about the Brownian bridge, in particular we
% will calculate $\Exp[B^2]$
%  \begin{ex}[Evans .37]
% \begin{align*}
% \Exp[B^2] =& \Exp\left[\left( (1-t) \int_0^t \frac{1}{1-s} dW \right)^2 \right]
% \\
% =&(1-t)^2 \Exp\left[ \int_0^t\left(  \frac{1}{1-s}\right)^2 \intd{s} \right]
% \\
% =&(1-t)^2 [ \frac{1}{1-t} - 1 ]
% \\
% =&(1-t) - (1-t)^2
% \end{align*}
% 
% Further note that  $\lim_{t\ra 1^-}\Exp[B^2] = 0$ and that $B$ is a martingale
% and so its first moment is always equal to its initial value, $0$. So indeed the 2nd
% moment is the variance and with the variance converging to zero, we can use
% Chebyshev's inequality to conclude that $\lim_{t\ra 1^-}B(t) = 0 $(a.s). This
% justifies the name of the Brownian bridge - it is a normally distributed RV.
% which is clamped to equal 0 at both $t= \{0,1\}$.
% \end{ex}

As an example we will discuss the Ornstein-Uhlenbeck process, which is the
basis for the models we will face later on.
\begin{ex}[O-U Process] Let $X$ follow:
\begin{equation*}
dX = \th (\a - X_t) dt + \b dW
\end{equation*}
with an initial condition, $X_0$, which may be an arbitrary distribution
independent of the Wiener Process,$W$.
We can solve this as follows:
\begin{align*}
dX =& \th (\a - X_t) dt + \b dW
\\
dX + \th X_t) dt=& \th \a dt + \b dW
\\
e^{\th t} dX + e^{\th t}\th X_t) dt=& \th e^{\th t}\a dt + \b e^{\th t} dW
\\
Xe^{\th t} - X_0 = & \int \th e^{\th t}\a dt +  \int \b e^{\th t} dW
\\
X_t =& e^{-\th t} X_0 + \a(1-e^{-\th t}) +  \frac{\b e^{-\th t}}{\sqrt{2\th}}
W(e^{2\th t}-1)
\end{align*}
which means that $X$ forgets its initial conditions exponentially fast and
converges to a normal random variable with mean $\a$ and variance
$\tfrac{\b^2}{2\th}$
\end{ex}

At the end of this sub-section, we state Ito's Lemma with multiple Wiener
processes and then for a multidimensional state.

\begin{thm}[Ito Lemma for $dW \in  \R^{m}$] Suppose $X$ satisfies the stochastic
differential $dX_t = F dt + G dW$, where $ F \in \R, G \in \R^{1\times m}$ and
$dW = \left(dW^{(i)}\right) \in  \R^{m}$ is a vector of independent Wiener
Processes. Take $u(x,t) \in C^{2,1}[ \R \times [0,T]]$. Let $D :=\left[{\sum_k
G_{k} G_{k} } \right] \in \R$

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u +  \di_{x} u \cdot F + 
  \di^2_{x} u \cdot D \right)
\intd{t} +
 \left(  \di_{x} u  \cdot  G \right) dW  
$$
\end{thm}

\begin{thm}[Ito Lemma for $X \in R^n$]
Suppose $X$ satisfies the stochastic differential $dX_t = F dt + G dW$, where $
F \in \R^n, G \in \R^{n\times m}$ and $dW = \left(dW^{(i)}\right) \in  \R^{ 
m}$ is a vector of independent Wiener Processes. Take $u(x,t) \in C^{2,1}[ \R^n
\times [0,T]]$. Let $D_{ij} := \tfrac{1}{2} \left[{\sum_k G_{ik} G_{jk} }
\right]_{ij} \in \R^{n\times n} $

Set $Y(t) = u(X,t)$
\\
then
$$
dY =  \left( \di_t u + \sum_i \di_{x_i} u \cdot f_i + 
\sum_{i,j} \di^2_{x_i x_j} u \cdot D_{ij} \right)
\intd{t} +
 \left(  \sum_{ij} \di_{x_i} u  \cdot G_{ij} dW^{(j)} 
\right)$$
\end{thm}

\subsection{Fokker-Planck and Kolmogorov's Backward Equations}
The Fokker-Planck equation and Kolmogorov's Backward equation describe the
forward (resp. backward)  evolution of the probability density of $X_t$

For the rest of this section we will work in $ \R^n$, i.e. $X$ will be an $n-$
dimensional vector that satisfies the stochastic differential 
\begin{equation}
dX_t = F(X,t) dt + G(X,t) dW,
\label{eq:generic_Ito_SDE_Rn}
\end{equation}
$F \in \R^n, G \in \R^{n \times m}$,
$dW = (dW^{(i)}) \in  \R^{  m}$ is an $m$-dimensional Brownian motion and we
write $D_{ij} :=
\tfrac{1}{2} 
\left[{\sum_k G_{ik} G_{jk} } \right]_{ij} \in \R^{n\times n} $.

% Let $$\F(x,t| y,s) = \Prob[X_t \preceq x| X_s = y] $$ be the forward transition
% distribution and

Let $$ \f(x,t| y,s) \intd{x} =  \Prob[X_t \in \intd{x} | X_s = y] $$ be the
transition probability density. Then $\f$ satisfies:
\begin{equation}
\di_t \f= -\sum_i \di_{x_i} \left[ F_i(x,t) \f(x,t) \right] 
+ 
\sum_{i,j}  \di^2_{x_i x_j} \left[ D_{ij}(x,t) \f(x,t) \right]
\label{eq:FP_pde}
\end{equation}
This is called the \emph{Fokker-Planck} or \emph{Forward Kolmogorov} equation.
It can also be seen as a continuity euquation or a cosevation of probability
equation. To this end define the probability current, $\p \in \R^n$ as:
$$
\phi_i =   F_i(x,t) \f(x,t) 
+ 
\sum_{j}  \di_{x_j} \left[ D_{ij}(x,t) \f(x,t)\right]
$$
then the Fokker-Planck equation, \cref{eq:FP_pde}, can be written as:
$$
\di_t \f= -\grad \cdot \phi
$$
which just says that the change in probability is the
difference between the flow in and the flow out.

Initial conditions for $\f$ are given by the distribution of $X_s$, if $s=0$
then this is the initial distribution of $X$. Boundary conditions (BCs) depend
on the domain of $X$ and what happens to $X$ once it hits its boundary. If the domain
is all of $\R^n$ then we only insist that $\lim_{|x| \ra \infty} \f = 0$. If the
domain has boundaries, then there are two common scenarios which we will also
encounter in the sequel:
\begin{enumerate}  
  \item absorbing BCs
  \item reflecting BCs
\end{enumerate}
At an absorbing boundary, the particle $X$ is removed and $\f=0$ there.
In this situation, we will not have conservation of probability and the integral
of $\f$ over $X$'s domain will monotonically decrease.

At a reflecting boundary, the particle $X$ bounces back into its domain and we
will have that the probability current $\p\cdot n = 0$ where $n$ is the outward
normal at the reflecting boundary.

We now consider $\f$ as functions of $y,s$, holding $x,t$ fixed, then
\begin{equation}
-\di_s \f= \sum_i  \left[F_i(y,s) \cdot \di_{y_i}\f(|y,s) \right] 
+ 
\sum_{i,j}   \left[ D_{ij}(x,t) \di^2_{y_i y_j}\f(|y,s) \right]
\label{eq:FP_backward_pde}
\end{equation} 
This is the \emph{Backward Kolmogorov} equation for $\f$. Note the minus sign
in front $\di_s \f$. A mnemonic for the signs of the Backward vs. Forward
equation is that as $t$ increases $\f$ diffuses and so $\di_t$ and $\di^2_{x}$
have the same sign, but as $s$ increases $\f$ anti-diffuses and they have
opposite signs.

The differential operator on the right-hand side of
\cref{eq:FP_backward_pde} occurs often in the study of SDEs and has its own
name.
\begin{defn}[Generator of an SDE] the Generator $A$ of $X$ is defined by:
$$
A[\psi(x)] = \lim_{t \searrow 0^+} \frac{\Exp[\psi(X_t)]  - \psi(x)}{t} ;  \quad X_0 =
x \in \R^n
$$
\end{defn} 
\begin{lemma} For an Ito SDE as in \cref{eq:generic_Ito_SDE_Rn} 
$$
A[\psi(x)] = \sum_i F_i(x,t) \cdot \di_{x_i} \psi + \sum_{i,j} D_{ij}(x,t)
\di^2_{x_i x_j}\psi $$
\end{lemma}

\subsection{Stopping Times}
\begin{defn}
Let $\Fil(t)$ be some filtration, a random variable $\t$ is called a
\emph{stopping time} if 
$$
\{\o : \t(\o) \leq t\} \in \Fil(t) \, \forall t  
$$
\end{defn}
The colloquial way of describing stopping times is at any time we know
whether $\t$ has occurred or not. For a counterexample, the time that a Wiener
Process achieves its maximum over some interval is not a stopping time, since at any
given time, we do not know if the maximum has occurred or not. The most common
example of a stopping time is the first hitting-time, which is the first time
$X_t$ leaves a set. 
\begin{thm}[First-Hitting time] Let $E \subset \R^n, E \neq
\phi$ be open or closed

then $$\t := \inf\{ t \geq 0 | X_t \in E\}$$ is a stopping time.
\end{thm}

The reason stopping times are very useful is that all the facts so far quoted
for Ito calculus using integrals $\int_0^T \intd{W}$ remain true if $T$ is
replaced by a stopping time $\t$.

>>>>

\usepackage{cancel}
\usepackage{cleveref}
% \crefname{algocf}{alg.}{algs.}
% \Crefname{algocf}{Algorithm}{Algorithms}

@@

\title{ISI Control in stochastic LIF models}
\author{Alexandre Iolov 
$<$\href{mailto:aiolo040@uottawa.ca}
		{aiolo040 at uottawa dot ca}$>$}

\date{\today}

@@

\abstract{Given a noisy and leaky integrate-and-fire neuronal model - we pose an
optimal control problem - how do we ensure a certain spike sequence. We then ask
whether and how can we do this online, in 'real-time'?}

@@

\section{Problem Formulation}
Given a noisy LIF neuronal model:
\begin{equation}
\begin{gathered}
dX_s = (\a(t) - \frac{X_s}{\tc} ) \intd{s} + \b \intd{W_s},
\\
X(0) = .0,
\\
X(\ts) = \xth \implies  
\begin{cases}
X(\ts^+) = .0 &  
\end{cases}
\end{gathered}
\label{eq:X_evolution_uo}
\end{equation}

So $X$ represents the membrane el. potential which tries to revert to $0$ with a
time-const of $\tc$, $dW$ is a Brownian motion increment with intensity $\b$.
$\a(t)$ is our external control and having last spiked at time $0$, the
potential hits $\xth$ at time
$t$, then we record the next 'spike' of duration $t$ and reset the potential to
$0$.

Suppose we want want to ensure that spikes occur exactly at the predetermined
times $\tns$. Then given the renewal nature of our spiking model, we can work
sequentially, focus only on the first of these intervals, of duration $\T$ and
then ask: how do we choose $\a(t)$ st. $\ts= \T$. This is not specific enough
since the probabilistic aspects of the problem ensure that we will not be able
to hit $\T$ precisely, unless we take extreme forms for $\a$, which will
naturally be disallowed in the sequel.

In the probabilistic context, there are several ways to specify to what we
mean by 'make $\ts = \T$', all of which result in optimal control problems.

The first is to ensure that we maximize the probability of $\T$. Let:
\begin{equation}
\begin{array}{lcll}
g_{\a}(\t) \intd{\t} &:=& \Prob[\ts \in [\t, \t + \intd{\t}] ] &
 \textrm{(probability density)} 
\\
G_{\a}(t) &:=& \Prob[\ts \leq t ] = \int_0^t g_{\phi}(\t) \intd{\t}
&
 \textrm{(cumulative distribution)}
\\
\G_{\a}(t) &:= & \Prob[\ts>t] = 1 - G_{\a}(t)
&
 \textrm{(survivor distribution)}
\end{array}
\label{eq:ISI_distribution_functions}
\end{equation}
be the various distributions for the spike time $\ts$. Then given $\T$, we want
to take:
\begin{equation}
\a(t) = \argmax_{\a(t)}  g_{\a}(\T)
\label{eq:OC_MLE_density} 
\end{equation}
or in a softer sense:
\begin{multline}
\a(t) = \argmax_{\a(t)}  [G_{\a}(\T+\delta T) - G_{\a}(\T-\delta T)]
\\ 
	  = \argmin_{\a(t)} [\G_{\a}(\T+\delta T) - \G_{\a}(\T-\delta T)]
\label{eq:OC_MLE_distribution} 
\end{multline}
for some small positive variation,  $\delta T$.

Alternatively, we can desire a least squares solution:
\begin{equation}
\a(t) = \argmin_{\a(t)} \{ \,(\Exp[\ts] - \T)^2  \, \}
\label{eq:OC_LS_mean}   
\end{equation}
or 
\begin{equation}
\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2] \,\}
\label{eq:OC_LS_variance}   
\end{equation}
i.e we can minimize the mean or the mean square; or alternatively we can
minimize both at the same time:
\begin{equation}
\a(t) = \argmin_{\a(t)} \{\, \Exp[(\ts- \T)^2  ] + m \Exp[(\ts- \T)^2 \,\}
\label{eq:OC_LS_mean-variance}   
\end{equation}
where $m$ sets our preference for mean-square minimization over minimization
just in the mean.

% Note that all three of
% \cref{eq:OC_LS_mean,eq:OC_LS_variance,eq:OC_LS_mean-variance} contain the $\Exp$
% operator and as such seemingly require us to know the distributions of $\ts$,
% i.e.\ $g_{\a}(\t)$, $G_{\a}(t)$ and $\G_{\a}(t)$ or do we\ldots?

% Since we have no hope of finding the density, $g_{\a}$, analytically, we will
% naturally resort to numerical methods both for calculating $g_{\a}(\t)$ and for
% solving the various optimal control problems.

So far we have not said anything about the constraints on the control. For now
we will just take simple bounded constraints:
\begin{equation}
\a(t) \in [\amin, \amax]
\label{eq:bound_constraints_alpha}
\end{equation}

\subsection{Dynamic Programing Basics}
We want to use Dynamic Programing to solve for the optimal control $\a(t)$.
Dynamic Programming has a well-developed theory for the control of both
diffusion- and jump- stochastic differential equations, see e.g.\
\cite{Hanson2007,Øksendal2005}. Here we just state the main result for
the diffusion case, with a non-controlled diffusion term.

Let the controlled dynamics be:
\begin{equation}
dX = a(x,t, \a) \intd{t} + b(x,t)\intd{W}; \quad X_0 = x_0
\end{equation} 
where $X$ is the 'state' and $\a$ is the 'control'.

Let the objective be:
\begin{equation}
J = \int_0^T L(s,x(t),\a(t)) \intd{s} + M(X_T)
\end{equation} 
Suppose we are minimizing, and set the value function, also called the
optimal cost-to-go, be:
\begin{equation}
v(x,t) = \inf_{\a(s>t)} \int_t^T L(s,x,\a) \intd{s} + M(X_T); \quad X_t = x
\end{equation}
then $v$ satisfies the following  backward parabolic PDE, the
Hamilton-Jacobi-Bellman (HJB) equation:
\begin{equation}
\di_t v(x,t) + \inf_{\a}  \{ L(t,x,a)  + a(x,t,\a)\di_x v + \tfrac{b^2(x,t)}{2}
\di^2_x v \}
\end{equation}
i.e.\ we tabulate the optimal value for any given $x$ in the state-space at any
given time $t$. With the terminal conditions:
\begin{equation}
v(x,T) = M(x)
\end{equation}

The boundary conditions on $v$ are quite context-specific and we will see how
they look for our problem below.

The optimal control is embedded in the evolution of $v$ and given $x,t$
and $v(x,t)$ we will have: 
\begin{equation}
\a(t) = \inf_{\a}  \{ L(t,x,a)  + a(x,t,\a)\di_x v + \tfrac{b^2(x,t)}{2}
\di^2_x v \}
\end{equation}

\subsection{The 'feedback' for our problem}
The key product of an application of Dynamic Programing is a feedback control
$\a(t) = \a(x(t))$, i.e.\ the current control depends on the current state (the
observe-react paradigm). However, in our particular problem, we cannot observe
the state.

We are not completely blind - at time $t$ we know whether $X$ has spiked
or not! So if $t > \T$, and we haven't spiked, we know we must put $\a =
\amax$. But for $t < \T$, we don't know where $X_t$ is.

To repeat - we are in some limbo state between fully observing the state and
knowing nothing about the state - we know whether the spike has occurred.

If we can fully observed the state, we will know what $X_t$ is; if we know
nothing about the state we only have its unconstrained transition density
$\rho(x,t)$ of $X_t$. However, in our case, we know the transition density,
$f(x,t)$, of $X_t$, conditioned on $X_{s <=t} < \xth$. The main difference
between the $\rho$ and $f$ - is that $\rho(x,t)$ gives a non-zero probability to
$X_t > \xth$, whereas we have the knowledge that this is impossible. And so
our actual optimal control law - the one that will be a candidate for
'real-world' implementation will be one that controls not $X_t$, but the
computable transition density $f(x,t)$.

% \section{Deterministic Control}
% \label{sec:deterministic_control}
% Let us first discuss the problem in the simpler, deterministic setting. We will
% assume that given $\T$ it is possible to get from $x_0$ to $\xth$ in time $t =
% \T$. Our problem then is:
% 
% Find $\a(t) \in [\amin, \amax]$ st. 
% \begin{equation}
% \begin{gathered}
% \dot{x} = \a(t) - x_t/\tc
% \\
% \begin{cases}
% x(.0) = .0 \\
% x(\T) = \xth \\
% x(t < \T) < \xth
% \end{cases}
% \end{gathered}
% \label{eq:X_evolution_deterministic}
% \end{equation}
% Ok, this problem is entirely ill-posed (there are very many solutions, $\a(t)$)
% and so to pick out one in particular we need to put some extra constraints on the control,
% $\a$.

\section{Maximum-Likelihood Control}
We will not dwell much on this here, except to make the following point. Suppose
the MLE distribution $g_{\a^*}$ is very slanted to the right, say it looks like:
\begin{equation}
g_{\a^*}(t) \propto 
\begin{cases}
0 & t < \T
\\
e^{-(t-T)}&  \ow
\end{cases}
\end{equation}
then the most likely $\ts$ will indeed be $\T$, but the bias etc. will be 'bad'.
However, since I am not an ML expert, I shouldn't comment much
more on whether this outcome is possible - maybe such a pathological situation
cannot occur and ML control is the right thing to do\ldots

\section{Minimization-in-the-Variance Control}
Here we discuss finding the control that minimizes \cref{eq:OC_LS_variance} plus
a running energy cost:
\begin{equation}
\a(t) = \argmin_{\a(t)} \Big\{ \,
\Exp\left[
\e \int_0^\ts  \a^2 \intd{s}
+  
(\ts - \T \big)^2 \right]  \, \Big\}
\label{eq:OC_LS_variance_energy}   
\end{equation}
where $\e$ measure how much weight we put on minimizing the energy expenditure
and if $\e \rightarrow 0$, then we don't care at all. This running cost is
similar to the regularization term $R(\x)$ in eq.\ 3 in \cite{Ahmadian2011}.

\subsection{Dynamic Programing Formulation}
Given $x,t$, let $\t$ be the (unknown) remaining time to spike. For example,
given $t=0, X=.0$, $\t = \ts$, the entire duration of the inter-spike interval.
Starting from a general $t, X_t=x$, our remaining-cost objective, $J[\a, x,t]$,
will be:
\begin{equation}
J[\a, x,t]  = \Big\{ \,
\Exp\left[
\e \int_t^\ts  \a^2 \intd{s}
+
(\t - (\T-t) \big)^2 |x,t
\right] \, \Big\}
\label{eq:OC_LS_mean_objective}   
\end{equation}
i.e.\ if time $t$ has elapsed without a spike, we now want to minimize
the difference between $\t$ and ($T-t$), given the current state $x$. For now we
assume we know $x$, later we will admit that we don't.
